Combining information from photometric and spectroscopic data:\\ propogating uncertainties in spectrophotometric calibration

### 2015-04-25 tex/ms.tex
move figures to end, change captions around

%figures
\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.pdf}
\newcommand{\excluster}{B192-G242}

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance
$s^2$.  For photometric data this ``jitter'' is often best described
as a fractional error, $(s\mu(\theta))^2$, corresponding for example
to an additional magnitude zero-point uncertainty.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + (s\,\mu_n(\theta))^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + (s\,\mu(\theta))^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  
In this formulation, if $\theta$ are the parameters of interest, the
jitter variance $s^2$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The
Gaussian Process will be a generalization of this modification.

@@

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to
the mean model $\mu$. In this formulation the the mean model $\mu$ will be effectively
multiplied by $e ^{f(\alpha)}$ 
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

@@

XXX Alternative XXX
To avoid making the logarithmic transform described above, which
distorts the errors, we can formulate the problem in a different way,
modifying Equation \label{eq:spectroscopicLF} above as 
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) * f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (s * \mu)^2]\,\delta_{nn'} +
\transpose{\mu} \, K_\alpha(\lambda_n - \lambda_{n'}) \, \mu
\quad ,
\end{eqnarray}
where the multiplication of $s$ and $K_\alpha$ by $\mu$ in the
construction of $C_{nn'}$ means that they are defined in
\emph{fractional} terms, and the uncertainties remain Gaussian in
linear flux space.  It remains to be seen whether this actually works.

@@

This separate treatment of mock and real data allows us to address two
separate issues.  The inference from mock data (with a realistic
calibration vector applied) provides a test of the methodology and the
effect of calibration uncertainties on the inferred parameters in the
case of a perfect mean model $mu(\theta)$.  The inference from real
data tests a combination of our methodology and the ability of the
population synthesis models to recover parameters inferred from
resolved star CMDs (i.e. some estimate of ground truth).

@@

\begin{figure}[h!]
%\includegraphics[width = 0.5 \textwidth]{figures/dfig_b192-g242_020.png}
\caption{The calibrated spectrum for \excluster, along with the
uncertainties (bottom panel).The right panel shows the
spectrophotometric calibration vector determined by
\citet{schiavon05}. \label{fig:ggc_data}}
\end{figure}

@@

\begin{figure}[h!]
%\includegraphics[width = 0.5 \textwidth]{figures/dfig_b192-g242_020.png}
\caption{A mock spectrum for a star cluster with $10^5 M_\odot$ total
  stellar mass, an age of  9 Gyr, solar
  metallicity, and $A_V=0.5$ mag. \label{fig:mock_data}}
\end{figure}

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/sfig_b192-g242_225_cal.png}
\caption{Spectra resulting from draws from the posterior parameter
space conditioned on the calibrated \excluster~ spectrum and
photometry.  Left: For each of 5 draws from the posterior we show from
top to bottom, three components of the model: the mean cluster
spectrum, the calibration vector, and the mean prediction of $\Delta$
for the Gaussian process.  Right: From top to bottom, the full model
({\it magenta}) compared to the observed, calibrated spectrum ({\it
grey}), the ratio of the observed to the modeled spectrum, and the
difference between the observed spectrum and the modeled spectrum
normalized by the uncertainty.
\label{fig:speconly_calibrated}}
\end{figure*}

@@

\subsection{Mock data}

In Figure \ref{fig:speconly_calibrated} we show the result of our
inference from the spectrum and a single photometric measurement (SDSS
$g$) for a perfectly calibrated mock spectrum. To demonstrate the
quality of the fit we show both the mock
input spectrum and predictions of the observed spectra that result
from several draws of the model parameters
$[\theta, \alpha]$ from posterior distribution.  We also show samples
of the inferred calibration vector,
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & e^{f(\alpha) + \tilde{\Delta}}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian process mean prediction for the
logarithmic residual between the data and the model.  For the
perfectly calibrated spectrum the true calibration vector (shown in
black) is one everywhere.  The spread of the inferred calibration
vectors approximately indicates our uncertainty in the spectrophotometric
calibration when no additional calibration information is available
and only a single photometric point is available.

Even in the case of a single photometric data point, several of the
physical parameters are well-constrained and only minimally biased,
e.g. age and metallicity.  These are the parameters that are
constrained well by the spectrum itself and are not strongly sensitive
to the overall shape of the spectral continuum.  Such parameters can
be reasonably well recovered by fitting continuum normalized spectra,
though hewre we have propogated the uncertainty in this normalization
into the uncertainties on the physical parameters.

@@

\subsection{Real Data: NGCXXX} 
We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.

@@

It is possible to reconstruct the inferred total multiplicative
calibration vector for each draw of the model parameters.  Rearranging
equation \ref{eq:spectroscopicLF} yields
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & e^{f(\alpha) + \tilde{\Delta}}
\quad ,
\end{eqnarray}
for the total effective multiplicative calibration, where
$\tilde{\Delta}$ is now the Gaussian process mean prediction for the
logarithmic residual between the data and the model. The inverse of
this ratio is shown in the left panel of Figure
\ref{fig:inferred_calib} as our inferred total multiplicative
calibration function; if the spectrophotometric calibration were
perfect and we had correctly inferred it, then we would expect this
vector to be constant as a function of wavelength.

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/pfig_b192-g242_225_cal.png}
\caption{ Model SEDs.  Top: The model photometric SED resulting from
draws from the posterior parameter space conditioned ({\it cyan}) on
the observed photometry ({\it magenta}) and spectra.  Bottom:
uncertainty normalized residuals of the models from the observations.
\label{fig:inferred_sed}}
\end{figure}

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/ptri_b192-g242_225_cal.png}
\caption{ Marginalized and joint PDFs of subset of the physical
parameters $\theta$ inferred from the spectophotometrically calibrated
spectrum of \excluster. The blue lines indicate the parameter values
that resulted from the powell minimization.
\label{fig:inferred_params}}
\end{figure}

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/cfig_b192-g242_225.png}
\caption{Left: Samples of the effective multiplicative calibration
vector for the spectrophotometrically calibrated spectrum of
\excluster~ ({\it cyan}) compared to that obtained from ancillary
photometric data ({\it magenta}).  
Right: The same, but for the uncalibrated data.
 \label{fig:inferred_calib}}
\end{figure*}

\begin{figure}[h!]
%\includegraphics[width =0.5\textwidth]{figures/B192_020_cratio.pdf}
\caption{Money plot.  Show the difference between the parameters
  inferred from the calibrated and uncalibrated spectra, for each
  cluster. Tjis is a placeholder, since it only shows B192 for fixed
  GP parameters. \label{fig:boom}}
\end{figure}

@@

We have demonstrated, using using both mock and real spectra of
galactic stellar clusters, that we can combine photometric and
spectroscopic data in a principled way to infer physical properties.
We do not resort to an arbitrary down-weighting or up-weighting of
certain data. Instead, we explicitly model our uncertain knowledge of
the spectral calibration and infer this calibration jointly with the
physical properties. For a number of applications this combination of
spectroscopy and photometry reduces the need for spectrophotometric
calibration entirely, such that the less expensive photometric
calibration is sufficient.  Additional calibration information, if it
exists, can be incorporated into informative priors on the calibration
parameters of the model, inform the construction of the calibration
model, or approximately remove gross features that are difficult to
model parameterically or via a Gaussian Process.

@@

Our strategy is easily extensible to more complicated models, which
include for example night sky-line emission in the spectroscopy (or
residuals from an intial sky-line subtraction), and uncertainties in
wavelength calibration. Furthermore, more complex physical models may
be considered for the mean object model, for example including SFH
parameters so as to model galaxies.  {\color{blue} DISCUSS MORE
COMPLICATED COVARIANCE KERNELS, MULTIPLE KERNELS, etc.  MODEL
CHECKING.}

@@

>>>>

%figures
\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.pdf}
\newcommand{\excluster}{NGC7089}

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance
$s^2$.  For photometric data this ``jitter'' is often best described
as a fractional error, $(\hat{s}\mu(\theta))^2$, corresponding for example
to an additional magnitude zero-point uncertainty.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  In this formulation, if $\theta$ are
the parameters of interest, the jitter variance $\hat{s}^2\equiv
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The Gaussian Process
will be a generalization of this modification.

@@

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to
the mean model $\mu$. In this formulation the the mean model $\mu$ will be effectively
multiplied by $e ^{f(\alpha)}$ 
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

@@

XXX Alternative XXX
To avoid making the logarithmic transform described above, which
distorts the errors, we can formulate the problem in a different way,
modifying Equation \label{eq:spectroscopicLF} above as 
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) * f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} * \mu)^2]\,\delta_{nn'} +
\transpose{\mu} \, K_\alpha(\lambda_n - \lambda_{n'}) \, \mu
\quad ,
\end{eqnarray}
where the multiplication of $s$ and $K_\alpha$ by $\mu$ in the
construction of $C_{nn'}$ means that they are defined in
\emph{fractional} terms, and the uncertainties remain Gaussian in
linear flux space.  It remains to be seen whether this actually works.

@@

This separate treatment of mock and real data allows us to address two
separate issues.  The inference from mock data (with a realistic
calibration vector applied) provides a test of the methodology and the
effect of calibration uncertainties on the inferred parameters in the
case of a perfect mean model $\mu(\theta)$.  The inference from real
data tests a combination of our methodology and the ability of the
population synthesis models to recover parameters inferred from
resolved star CMDs (i.e. some estimate of ground truth).

@@

At each iteration of the sampler the photometric and spectroscopic
likelihood are computed, multiplied together, and multiplied by the
the prior probability of that parameter position to obtain an estimate
of the posterior probability for those parameters given the data.

@@

@@

\subsection{Mocks}
We first present the results of our inference from a spectrum where we
assume the calibration is known perfectly.  That is, we fix the
calibration parameters $\alpha$ to the values used to construct the
mock data.  This sets a baseline for the kind of posterior PDFs that
could be obtained for the physical parameters given the
signal-to-noise of the mock data, perfect models, and perfectly known
calibration.

{\color{blue}BDJ Describe results from the ideal case.}

Next we will \emph{not} assume that the calibration is perfectly known
and we will therefore allow the calibration parameters $\alpha$ in our
model to vary.  In Figure
\ref{fig:speconly_calibrated} we show the result of our inference from
the spectrum and a single photometric measurement (SDSS $g$) for a
perfectly calibrated mock spectrum. To demonstrate the quality of the
fit we show both the mock input spectrum and predictions of the
observed spectra that result from several draws of the model
parameters $[\theta, \alpha]$ from posterior distribution.  We also
show samples of the inferred calibration vector,
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & e^{f(\alpha) + \tilde{\Delta}}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian Process mean prediction for the
logarithmic residual between the data and the model.  For the
perfectly calibrated spectrum the true calibration vector (shown in
black) is one everywhere.  The spread of the inferred calibration
vectors approximately indicates our uncertainty in the spectrophotometric
calibration when no additional calibration information is available
and only a single photometric point is available.

Even in the case of a single photometric data point, many of the
physical parameters are well-constrained and only minimally biased,
e.g. age and metallicity.  These are the parameters that are
constrained well by the spectrum itself and are not strongly sensitive
to the overall shape of the spectral continuum.  Such parameters can
be reasonably well recovered by fitting continuum normalized spectra,
though here we have propogated the uncertainty in this normalization
into the uncertainties on the physical parameters.

@@

\subsection{Real Data: NGCXXXX} 
We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.

@@

It is possible to reconstruct the inferred total multiplicative
calibration vector for each draw of the model parameters.  Rearranging
equation \ref{eq:spectroscopicLF} yields
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & e^{f(\alpha) + \tilde{\Delta}}
\quad ,
\end{eqnarray}
for the total effective multiplicative calibration, where
$\tilde{\Delta}$ is now the Gaussian Process mean prediction for the
logarithmic residual between the data and the model. The inverse of
this ratio is shown in the left panel of Figure
\ref{fig:inferred_calib} as our inferred total multiplicative
calibration function; if the spectrophotometric calibration were
perfect and we had correctly inferred it, then we would expect this
vector to be constant as a function of wavelength.

@@

We have demonstrated, using using both mock and real spectra of
galactic stellar clusters, that we can combine photometric and
spectroscopic data in a principled way to infer physical properties.
We do not resort to an arbitrary down-weighting or up-weighting of
certain data. Instead, we explicitly model our uncertain knowledge of
the spectral calibration and infer this calibration jointly with the
physical properties. For a number of applications this combination of
spectroscopy and photometry reduces the need for spectrophotometric
calibration entirely, such that the less expensive photometric
calibration is sufficient.  Additional calibration information, if it
exists, can be incorporated into informative priors on the calibration
parameters of the model, inform the construction of the calibration
model (including the form of the covariance kernel), or approximately
remove gross features that are difficult to model parametrically or
via a Gaussian Process.

@@

Our strategy is easily extensible to more complicated models, which
include for example night sky-line emission in the spectroscopy (or
residuals from an intial sky-line subtraction) and uncertainties in
wavelength calibration. Furthermore, more complex physical models may
be considered for the mean object model, for example including SFH
parameters so as to model galaxies.  {\color{blue} DISCUSS MORE
COMPLICATED COVARIANCE KERNELS, MULTIPLE KERNELS, etc.  MODEL
CHECKING.}

@@

\input{figures.tex}
